Quantum Interval-Valued Probability: Contextuality and the Born Rule

### 2017-09-30 prop_letter.tex
prop_letter.tex \section{Gleason} Keep reconstructing...

In Sec.~\ref{sec:Interval-Uncertainty}, we discussed how states
consistent with a QIVPM~$\bar{\mu}$ on the projectors in a subspace
of event~$\events'$. In particular, if $\events'$ commutes, $\bar{\mu}$
restricting on $\events'$ is essentially a classical IVPM, and Thm.~\ref{thm:Shapley}
claims that there is always a state~$\rho$ consistent with $\bar{\mu}$.
In this section, we want to shift our attention to consider states
consistent with $\bar{\mu}$ on every event, $\coreBorn\left(\bar{\mu},\events\right)$,
so that these states could characterize $\bar{\mu}$ comprehensively.
When the intervals are infinitely precise, i.e., $\mathscr{I}_{\infty}=\set{\left[x,x\right]}{x\in\left[0,1\right]}$,
QIVPMs become to quantum probability measures~\citep{HOSTunpublished},
and Gleason's theorem claims that there is always a unique state consistent
with $\bar{\mu}:\events\rightarrow\mathscr{I}_{\infty}$~\citep{gleason1957,Redhead1987-REDINA,peres1995quantum}.
However, if we consider general QIVPMs mapping to an \emph{arbitrary}
set of intervals, there is a QIVPM that no state can be consistent
with it.

@@

\noindent To prove the theorem, we first need to construct a QIVPM,
and then verify that no states consistent with it on all projectors.
Given an orthonormal basis $\left\{ \ket{0},\ket{1},\ket{2}\right\} $,
let $\mathscr{I}_{0}=\left\{ \necess,\imposs,\unknown\right\} $,
$\ket{\ps}=\left(\ket{0}+\ket{1}\right)/\sqrt{2}$, $\ket{\ps'}=\left(\ket{0}+\ket{2}\right)/\sqrt{2}$.
The function~$\bar{\mu}:\events\rightarrow\mathscr{I}_{0}$ defined
in Table~\ref{tab:non-Born-QIVPM} can be verified as a QIVPM straightforwardly.
\begin{table}
\caption{\label{tab:non-Born-QIVPM}Define a non-Born QIVPM~$\bar{\mu}:\events\rightarrow\mathscr{I}_{0}$
on the three-dimensional Hilbert space. The projectors mapping to
the same interval is listed in the column labeled by $P$ and separated
by commas.}

@@

In other words, if the measurement intervals are too imprecise, the
results we get might contradict to each other, and exclude all possible
states consistent with the measurement results. There are many ways
to avoid these kind of too imprecise intervals~\citep{HOSTunpublished}.
Following, we consider a concrete family of intervals and QIVPMs~$\bar{\mu}$
such that if we monotonically shrink the length of intervals, the
states consistent with $\bar{\mu}$ will be closer and closer to $\rho$
which inducing $\bar{\mu}$.
\begin{equation}
\begin{aligned}\mathscr{I}_{n}= & \set{\left[\tfrac{k-1}{2n},\tfrac{k+1}{2n}\right]\cap\left[0,1\right]}{k=0,\ldots,2n}\\
 & \cup\{\imposs,\necess\}
\end{aligned}
\end{equation}
which are overlapping intervals with length~$1/n$ except for the
intervals with length~$\left(2n\right)^{-1}$ at each end. As shown
in Table~\ref{tab:I_n}, we select the correct interval from these
overlapping intervals~$[\ell,r]$, and define $\textrm{Int}\left([\ell,r]\right)$,
the non-overlapping interior part of an interval.
\begin{table}
\caption{\label{tab:I_n}Intervals in $\mathscr{I}_{n}$, their lengths, and
their interiors, where $2\mid k$ means $k$ is even, and $2\nmid k$
means $k$ is odd. Although intervals~$[\ell,r]$ are overlapping,
their interior~$\textrm{Int}\left([\ell,r]\right)$ are non-overlapping.
See Table~\ref{tab:I_2} for a concrete example when $n=2$.}

\begin{tabular}{cccc}
\toprule 
\addlinespace
$k$ & $[\ell,r]$ & $r-\ell$ & $\textrm{Int}\left([\ell,r]\right)$\tabularnewline\addlinespace
\midrule
\midrule 
\addlinespace
 & $\imposs$ & $0$ & $\imposs$\tabularnewline\addlinespace
\midrule 
\addlinespace
$0$ & $\left[0,\tfrac{1}{2n}\right]$ & $\tfrac{1}{2n}$ & $\left(0,\tfrac{0.5}{2n}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
\begin{tabular}{c}
$0<k<2n$\tabularnewline
and $2\nmid k$\tabularnewline
\end{tabular} & $\left[\tfrac{k-1}{2n},\tfrac{k+1}{2n}\right]$ & $\tfrac{1}{n}$ & $\left(\tfrac{k-0.5}{2n},\tfrac{k+0.5}{2n}\right)$\tabularnewline\addlinespace
\midrule 
\addlinespace
\begin{tabular}{c}
$0<k<2n$\tabularnewline
and $2\mid k$\tabularnewline
\end{tabular} & \multirow{1}{*}{$\left[\tfrac{k-1}{2n},\tfrac{k+1}{2n}\right]$} & $\tfrac{1}{n}$ & $\left[\tfrac{k-0.5}{2n},\tfrac{k+0.5}{2n}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$2n$ & $\left[\tfrac{2n-1}{2n},1\right]$ & $\tfrac{1}{2n}$ & $\left[\tfrac{2n-0.5}{2n},1\right)$\tabularnewline\addlinespace
\midrule 
\addlinespace
 & $\necess$ & $0$ & $\necess$\tabularnewline\addlinespace
\bottomrule
\end{tabular}
\end{table}
Then, given a real-valued probability~$x$, we will select the interval~$\iota_{n}\left(x\right)=[\ell,r]$
if $x\in\textrm{Int}\left([\ell,r]\right)$.

After we have the rule to replace a real number by an interval, we
can induce a QIVPM from a state according to the following theorem:

\begin{thm}[Finite-precision Extension of the Born rule]\label{thm:Finite-precision-Born}For
any state $\rho$ on a three-dimensional Hilbert space, a function~$\left[\mu_{\rho}^{\mathrm{B}}\right]_{n}:\events\rightarrow\mathscr{I}_{n}$
defined by $\left[\mu_{\rho}^{\mathrm{B}}\right]_{n}\left(P\right)=\iota_{n}\left(\mu_{\rho}^{\mathrm{B}}\left(P\right)\right)$
is a QIVPM.\end{thm}

\noindent To prove that $\left[\mu_{\rho}^{\mathrm{B}}\right]_{n}$
is a QIVPM, it is sufficient to check Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion}).
Given $x_{0}$, $x_{1}$, and $x_{0}+x_{1}\in\left[0,1\right]$, it
is easy to verify that
\begin{equation}
\iota_{n}\left(x_{0}+x_{1}\right)\subseteq\iota_{n}\left(x_{0}\right)+\iota_{n}\left(x_{1}\right)\,.\label{eq:iota-add}
\end{equation}
This implies Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion})
when $P_{0}$ and $P_{1}$ are orthogonal. When $P_{0}$ and $P_{1}$
are not orthogonal on a three-dimensional Hilbert space, we have two
possibilities:
\begin{itemize}
\item Their complements are orthogonal so that we can apply Eq.~(\ref{eq:iota-add})
on their complements.
\item One of them projects onto a subspace of what the other one projects
onto.
\end{itemize}
\noindent Both cases deduce Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion})
easily so that $\left[\mu_{\rho}^{\mathrm{B}}\right]_{n}$ is indeed
a QIVPM.

The collection of states consistent $\left[\mu_{\rho}^{\mathrm{B}}\right]_{n}$
gets closer and closer to $\rho$ as the intervals get narrower and
narrower, eventually collapsing to $\rho$ when the intervals become
infinitely precise real numbers covering the real line. To state this
result, we need to introduce a notion of distance between two states~$\rho$
and $\rho'$ as $\left\Vert \rho-\rho'\right\Vert =\max_{\ip{\psi}{\psi}=1}\left|\melem{\psi}{\rho-\rho'}{\psi}\right|$~\citep{544199}.
By replacing the QIVPM~$\bar{\mu}$ in Thm.~\ref{cor:Gleason's}
by a Born QIVPM~$\left[\mu_{\rho}^{\mathrm{B}}\right]_{n}$, we obtain
the following theorem that estimates the states consistent with $\left[\mu_{\rho}^{\mathrm{B}}\right]_{n}$.

@@

@@

\noindent If a state~$\rho'$ is consistent with $\left[\mu_{\rho}^{\mathrm{B}}\right]_{n}$,
the difference between their Born quantum probabilities on any 1-dimensioal
projector~$\proj{\psi}$, $\left|\mu_{\rho}^{\mathrm{B}}\left(\proj{\psi}\right)-\mu_{\rho'}^{\mathrm{B}}\left(\proj{\psi}\right)\right|$,
must be bounded by $1/n$ which implies $\left\Vert \rho-\rho'\right\Vert \le\frac{1}{n}$
by definition.

@@

In the limiting case when the intervals are infinitely precise, i.e.,
$\mathscr{I}_{\infty}=\set{\left[x,x\right]}{x\in\left[0,1\right]}$,
we can formulate Gleason's theorem in our framework as follows.

>>>>

In Sec.~\ref{sec:Interval-Uncertainty}, we discussed how states
consistent with a QIVPM~$\bar{\mu}$ on the projectors in a subspace
of event~$\events'$. In particular, if $\events'$ commutes, $\bar{\mu}$
restricting on $\events'$ is essentially a classical IVPM, and Thm.~\ref{thm:Shapley}
claims that there is always a state~$\rho$ consistent with $\bar{\mu}$.
In this section, we want to shift our attention to consider states
consistent with $\bar{\mu}$ on every event, $\coreBorn\left(\bar{\mu},\events\right)$,
so that these states could characterize $\bar{\mu}$ comprehensively.
When the intervals are infinitely precise, i.e., $\mathscr{I}_{\infty}=\set{\left[x,x\right]}{x\in\left[0,1\right]}$,
a QIVPM~$\bar{\mu}:\events\rightarrow\mathscr{I}_{\infty}$ becomes
quantum probability measure~\citep{HOSTunpublished}, and Gleason's
theorem claims that there is always a unique state consistent with
$\bar{\mu}$~\citep{gleason1957,Redhead1987-REDINA,peres1995quantum}.
However, if we consider general QIVPMs mapping to an \emph{arbitrary}
set of intervals, there is a QIVPM that no state can be consistent
with it.

@@

\noindent To prove the theorem, we first need to construct a QIVPM,
and then verify that no states consistent with it on all projectors.
Given an orthonormal basis $\left\{ \ket{0},\ket{1},\ket{2}\right\} $,
let $\mathscr{I}_{0}=\left\{ \necess,\imposs,\unknown\right\} $,
$\ket{\ps}=\left(\ket{0}+\ket{1}\right)/\sqrt{2}$, and $\ket{\ps'}=\left(\ket{0}+\ket{2}\right)/\sqrt{2}$.
The function~$\bar{\mu}:\events\rightarrow\mathscr{I}_{0}$ defined
in Table~\ref{tab:non-Born-QIVPM} can be verified as a QIVPM straightforwardly.
\begin{table}
\caption{\label{tab:non-Born-QIVPM}Define a QIVPM~$\bar{\mu}:\events\rightarrow\mathscr{I}_{0}$
on the three-dimensional Hilbert space. The projectors mapping to
the same interval is listed in the column labeled by $P$ and separated
by commas.}

@@

In other words, if the measurement intervals are too imprecise, the
results we get might contradict to each other, and exclude all possible
states consistent with the measurement results. To avoid this problem,
instead of proposing an abstract framework~\citep{HOSTunpublished},
we start from a concrete family of intervals as follow:
\begin{equation}
\begin{aligned}\mathscr{I}_{n}= & \set{\left[\tfrac{k-1}{2n},\tfrac{k+1}{2n}\right]\cap\left[0,1\right]}{k=0,\ldots,2n}\\
 & \cup\{\imposs,\necess\}
\end{aligned}
\end{equation}
which are overlapping intervals with length~$1/n$ except for the
intervals with length~$\left(2n\right)^{-1}$ at each end. As shown
in Table~\ref{tab:I_n}, for each interval~$[\ell,r]$, we define
$\textrm{Int}\left([\ell,r]\right)$, the non-overlapping interior
part of $[\ell,r]$.
\begin{table}
\caption{\label{tab:I_n}Intervals in $\mathscr{I}_{n}$, their lengths, and
their interiors, where $2\mid k$ means $k$ is even, and $2\nmid k$
means $k$ is odd. Although intervals~$[\ell,r]$ are overlapping,
their interior~$\textrm{Int}\left([\ell,r]\right)$ are non-overlapping.}

\begin{tabular}{cccc}
\toprule 
\addlinespace
$k$ & $[\ell,r]$ & $r-\ell$ & $\textrm{Int}\left([\ell,r]\right)$\tabularnewline\addlinespace
\midrule
\midrule 
\addlinespace
 & $\imposs$ & $0$ & $\imposs$\tabularnewline\addlinespace
\midrule 
\addlinespace
$0$ & $\left[0,\tfrac{1}{2n}\right]$ & $\tfrac{1}{2n}$ & $\left(0,\tfrac{0.5}{2n}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
\begin{tabular}{c}
$0<k<2n$\tabularnewline
and $2\nmid k$\tabularnewline
\end{tabular} & $\left[\tfrac{k-1}{2n},\tfrac{k+1}{2n}\right]$ & $\tfrac{1}{n}$ & $\left(\tfrac{k-0.5}{2n},\tfrac{k+0.5}{2n}\right)$\tabularnewline\addlinespace
\midrule 
\addlinespace
\begin{tabular}{c}
$0<k<2n$\tabularnewline
and $2\mid k$\tabularnewline
\end{tabular} & \multirow{1}{*}{$\left[\tfrac{k-1}{2n},\tfrac{k+1}{2n}\right]$} & $\tfrac{1}{n}$ & $\left[\tfrac{k-0.5}{2n},\tfrac{k+0.5}{2n}\right]$\tabularnewline\addlinespace
\midrule 
\addlinespace
$2n$ & $\left[\tfrac{2n-1}{2n},1\right]$ & $\tfrac{1}{2n}$ & $\left[\tfrac{2n-0.5}{2n},1\right)$\tabularnewline\addlinespace
\midrule 
\addlinespace
 & $\necess$ & $0$ & $\necess$\tabularnewline\addlinespace
\bottomrule
\end{tabular}
\end{table}
Since the interior part of intervals are non-overlapping, we can define
a function~$\iota_{n}:\left[0,1\right]\rightarrow\mathscr{I}_{n}$
by selecting the interval~$\iota_{n}\left(x\right)=[\ell,r]$ if
$x\in\textrm{Int}\left([\ell,r]\right)$.

After we have the way mapping a real-valued probability to an interval-valued
one, we can induce a QIVPM from a state according to the following
theorem:

\begin{thm}[Finite-precision Extension of the Born rule]\label{thm:Finite-precision-Born}For
any state $\rho$ on a three-dimensional Hilbert space, a function~$\left[\mu_{\rho}^{\mathrm{B}}\right]_{n}:\events\rightarrow\mathscr{I}_{n}$
defined by $\left[\mu_{\rho}^{\mathrm{B}}\right]_{n}\left(P\right)=\iota_{n}\left(\mu_{\rho}^{\mathrm{B}}\left(P\right)\right)$
is a QIVPM, called the Born QIVPM.\end{thm}

\noindent To prove that $\left[\mu_{\rho}^{\mathrm{B}}\right]_{n}$
is a QIVPM, it is sufficient to check Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion}).
Given $x_{0}$, $x_{1}$, and $x_{0}+x_{1}\in\left[0,1\right]$,
\begin{equation}
\iota_{n}\left(x_{0}+x_{1}\right)\subseteq\iota_{n}\left(x_{0}\right)+\iota_{n}\left(x_{1}\right)\label{eq:iota-add}
\end{equation}
implies Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion})
when $P_{0}$ and $P_{1}$ are orthogonal. When $P_{0}$ and $P_{1}$
are not orthogonal on a three-dimensional Hilbert space, we have two
possibilities:
\begin{itemize}
\item Their complements are orthogonal so that we can apply Eq.~(\ref{eq:iota-add})
on their complements.
\item One of them projects onto a subspace of what the other one projects
onto.
\end{itemize}
\noindent Both cases deduce Eq.~(\ref{eq:QuantumInterval-valuedProbability-Inclusion})
easily so that $\left[\mu_{\rho}^{\mathrm{B}}\right]_{n}$ is indeed
a QIVPM.

Because $\rho\in\coreBorn\left(\left[\mu_{\rho}^{\mathrm{B}}\right]_{n},\events\right)$,
there is always a state consistent with $\left[\mu_{\rho}^{\mathrm{B}}\right]_{n}$.
Moreover by introducing a notion of distance between two states~$\rho$
and $\rho'$ as $\left\Vert \rho-\rho'\right\Vert =\max_{\ip{\psi}{\psi}=1}\left|\melem{\psi}{\rho-\rho'}{\psi}\right|$~\citep{544199},
the collection of states consistent with $\left[\mu_{\rho}^{\mathrm{B}}\right]_{n}$
gets closer and closer to $\rho$ as the intervals get narrower and
narrower, eventually collapsing to $\rho$ when the intervals become
infinitely precise real numbers covering the real line.

@@

\noindent Given state~$\rho'$ consistent with $\left[\mu_{\rho}^{\mathrm{B}}\right]_{n}$,
the difference between their Born quantum probabilities on any 1-dimensioal
projector~$\proj{\psi}$ is $\left|\mu_{\rho}^{\mathrm{B}}\left(\proj{\psi}\right)-\mu_{\rho'}^{\mathrm{B}}\left(\proj{\psi}\right)\right|$,
and must be bounded by $1/n$ which implies $\left\Vert \rho-\rho'\right\Vert \le\frac{1}{n}$
by definition.

@@

In the limiting case when the intervals are infinitely precise, the
original Gleason theorem is recovered, and can be formulated in our
framework as follows.

@@
